---
title: "Introduction to the BiocAzul package"
author:
- name: Marcel Ramos
  affilitation: CUNY Graduate School of Public Health and Health Policy
  email: marcel.ramos@sph.cuny.edu
date: "`r format(Sys.time(), '%B %d, %Y')`"
package: BiocAzul
output:
  BiocStyle::html_document
abstract: |
  The BiocAzul package provides an interface to the Azul API, which is a service
  that allows users to search and access data from the Human Cell Atlas (HCA)
  and the AnVIL Data Explorer. This vignette introduces the BiocAzul package,
  demonstrating how to use it to query data and integrate it with Terra
  workspaces for further analysis.
vignette: |
  %\VignetteIndexEntry{Introduction to the BiocAzul package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# BiocAzul

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    cache = TRUE
)
```

# Installation

Install the development version of the `BiocAzul` package from GitHub using the
following:

```{r install, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Bioconductor/BiocAzul")
```

# Package Loading

```{r load-packages, message=FALSE, eval=TRUE, cache=FALSE}
library(BiocAzul)
```

# Introduction

The `BiocAzul` package provides an interface to the Azul API, which is used
to index data from the Human Cell Atlas (HCA) and the AnVIL Data Explorer.
Azul provides a convenient query interface for searching and retrieving data
from these projects.

# Basic Usage

To get started, create an `Azul` service object. By default, it connects to the
Human Cell Atlas service.

```{r azul-hca-init}
hca <- Azul()
hca
```

## Connecting to the AnVIL Data Explorer

To connect to the AnVIL Data Explorer instead, specify the provider when
creating the `Azul` object.

```{r azul-anvil}
anvil <- Azul(provider = "anvil")
anvil
```

Note that the `host` field in the objects output changes to reflect the AnVIL
Data Explorer service.

## Listing Catalogs

Azul organizes data into catalogs. You can list the available catalogs using
`listCatalogs()`.

```{r list-catalogs}
listCatalogs(hca)
```

## Exploring Projects

To get a quick overview of the projects in a catalog, use `projectTable()`.
This returns a `tibble` with project names and their corresponding IDs.

```{r project-table}
projects <- projectTable(hca, catalog = "dcp57")
head(projects)
```

## Exploring Facets

Azul data is organized by facets, which are attributes you can use to filter
and group data. You can list the available facets for a catalog using
`availableFacets()`.

```{r available-facets}
facets <- availableFacets(hca, catalog = "dcp57")
head(facets)
```

You can also get a summary of values for a specific facet using `facetTable()`.

```{r facet-table}
facetTable(hca, facet = "genusSpecies", catalog = "dcp57")
```

# Filtering and Queries

The `makeFilter()` function provides a convenient way to create filters for
querying the Azul API. It uses a formula-based syntax to define the filter
criteria.

```{r make-filter}
filter <- makeFilter(
    ~  specimenOrgan == "brain" &
        genusSpecies == "Mus musculus" &
        fileFormat == "h5"
)
filter
```

This filter can be used in other functions that interact with the Azul API.

# Integration with Terra

One of the main features of `BiocAzul` is the ability to import data directly
into a Terra workspace. This is done using the `importToTerra()` function.

Note: This step requires a Terra workspace and appropriate permissions. The
following code is for demonstration purposes and is not executed in this
vignette.

```{r import-to-terra, eval=FALSE}
importToTerra(
    hca,
    namespace = "your-terra-namespace",
    name = "your-terra-workspace",
    catalog = "dcp57",
    filters = filter
)
```

This function will create a manifest based on the filters, initiate an import
job in Terra, and poll for its completion.

# Session Information

<details>
  <summary>Click to see session information</summary>

```{r session-info}
sessionInfo()
```

</details>
